```{r}
#loading packages
library(dplyr)
library(tidyverse)
library(ggplot2)
library(sjPlot)
```


```{r}
view(cleaned_data)
```

#Data visualization

*Distribution of scores by year Plot*
```{r}
 ggplot(train_data, aes(x = factor(Year), y = Scores, fill = factor(Year))) +
  geom_violin(color = "black", alpha = 0.7, trim = FALSE) +
  geom_boxplot(width = 0.1,fill="white")+
  scale_fill_manual(values = c("brown","steelblue"))+
  ggtitle("Distribution of Scores by Year") +
  xlab("Year") + 
  ylab("Scores") +
  theme_minimal()
```

Here, both years distributions are roughly bell-shaped that is close to normal distribution.
In both years, The center or mean of the distribution is around 500.
Also, there's a symmetrical shape aroung the cluster indicating no major shift in average or spread of scores.

*Distribution of Scores by Year and Gender Plot*
```{r}
ggplot(cleaned_data, aes(x = factor(Year), y = Scores,fill = factor(Year))) +
  geom_violin(color = "black", alpha = 0.7, trim = FALSE) +
  geom_boxplot(width = 0.1, fill="white",outlier.shape = NA, alpha = 0.5) +
  scale_fill_manual(values = c("lightpink", "steelblue")) +
  ggtitle("Distribution of Scores by Year and Gender") +
  xlab("Year & Gender") +
  ylab("Math Scores") +
  theme_minimal() +
  theme(legend.title = element_blank())
(legend.title = element_blank())
```

*Plot of Parent Education vs Scores*
```{r}
cleaned_data$Parent_Educ<- factor(cleaned_data$Parent_Educ,
                                  levels=c("less than ISCED2","ISCED2","ISCED3",
                                           "ISCED4","ISCED6","ISCED7 or 8"))
cleaned_data$Year<-factor(cleaned_data$Year,levels=c("2015","2019"),ordered=TRUE)
ggplot(cleaned_data, aes(x = Parent_Educ, y = Scores, fill = Parent_Educ))+
  geom_boxplot(alpha = 0.6) + 
  ggtitle('Parent Education vs Scores') +
  facet_wrap(~Year)
  scale_fill_brewer(palette = "Set3")
```

*Plot of Parent Support vs Scores*
```{r}
ggplot(cleaned_data, aes(x = Parent_Support, y = Scores, fill = factor(Year))) + 
  geom_boxplot(alpha=0.6) + 
  ggtitle('Parent Support vs Scores')+
  facet_wrap(~Year)
  scale_fill_brewer(palette = "Set3")
```
*Computer Access vs Score based on Year*                   
```{r}
ggplot(cleaned_data, aes(x = Year, y = Scores, fill = Computer_Access)) +
  geom_boxplot(color = "black", alpha = 0.6) + 
  ggtitle('Computer Access vs Score based on Year') +
  scale_fill_manual(values = c("lightgreen", "red")) +
  facet_wrap(~Gender)
  theme_minimal()
```

*Distribution of Scores by Location and Gender*
```{r}
cleaned_data$Location<- factor(cleaned_data$Location,
                                  levels=c("Urban","Urban_1","Suburban","Suburban_1","Large Town/City","Small Town/Village","Remote Rural"),
                                          ordered= TRUE)
ggplot(cleaned_data, aes(x = Gender, y = Scores, fill = Location)) + 
  geom_violin(color = "black", alpha = 0.6) + 
  ggtitle('Distribution of Scores by Location and Gender') + 
   scale_fill_manual(values=c("navyblue","red","brown","orange","yellow","lightgreen","lightblue")) 
+facet_wrap(~Year)
```
**Interpretation**

Overall, all locations show a similar trend of having a center of distribution around 500-600.Urban locations may have slight wider violins. There is also no major differences between both the genders.Both the years shows similar shapes for all locations.
The width of the violins shows the density of scores and The overall variability (or spread) of scores is shown by the vertical extent of the violin.

*Plot of Distribution of Scores by Book Availability and Gender*
```{r}
cleaned_data$Book_Availability = factor(cleaned_data$Book_Availability,
                                        levels=c("Few","Moderate",
                                                  "Good","Very Good",
                                                           "Excellent"),
                                                            ordered=TRUE)
ggplot(cleaned_data, aes(x = Year, y = Scores, fill = Book_Availability)) + 
  geom_violin(color = "black", alpha = 0.6) + 
  ggtitle('Distribution of Scores by Book Availability and Gender') + 
  scale_fill_manual(values=c("lightblue","lightgreen","orange","brown","navyblue")) + 
  facet_wrap(~ Gender)
```
*Distribution of Scores by Teacher Shortage Based on Years*
```{r}
cleaned_data$Teacher_Shortage =factor(cleaned_data$Teacher_Shortage,
                                      levels=c("None","Low","Moderate",
                                                       "High"),ordered=TRUE)
ggplot(cleaned_data, aes(x = Year, y = Scores, fill = Teacher_Shortage)) + 
  geom_violin(color = "black", alpha = 0.6) + 
  ggtitle('Distribution of Scores by Teacher Shortage Based on Years') + 
  scale_fill_brewer(palette = "Set3")+
  facet_wrap(~Gender)
```
*Distribution of Scores by Percent Affluent Based on Years*
```{r}
cleaned_data$Percent_Affluent =factor(cleaned_data$Percent_Affluent,
                                      levels=c("Low","Moderate",
                                              "High","Excellent"),ordered=TRUE)
ggplot(cleaned_data, aes(x = Gender, y = Scores, fill = Percent_Affluent)) + 
  geom_violin(alpha = 0.6) + 
  ggtitle('Distribution of Scores by Percent Affluent Based on Years') + 
  scale_fill_brewer(palette = "Set1")+
  facet_wrap(~Year)
```
*Distribution of Scores by Percent Disadvantaged Based on Years*
```{r}
cleaned_data$Percent_Disadvantaged=factor(cleaned_data$Percent_Disadvantaged,
                                      levels=c("Low","Moderate",
                                              "High","Severe"),ordered=TRUE)
ggplot(cleaned_data, aes(x = Year, y = Scores, fill = Percent_Disadvantaged)) + 
  geom_violin(alpha = 0.6) + 
  ggtitle('Distribution of Scores by Percent Disadvantaged Based on Years') + 
  scale_fill_brewer(palette = "Set1")+
  facet_wrap(~Gender)
```

#Combined Box Plot by School ID     
```{r}

school_means <- cleaned_data %>%
  group_by(School_Id) %>%
  summarise(Mean_Score = mean(Scores, na.rm = TRUE)) %>%
  arrange(Mean_Score)

top_bottom_schools <- bind_rows(
  slice_head(school_means, n = 5),
  slice_tail(school_means, n = 5)
)

filtered_data <- cleaned_data %>%
  filter(School_Id %in% top_bottom_schools$School_Id)


filtered_data$School_Id <- as.factor(filtered_data$School_Id)


ggplot(filtered_data, aes(x = reorder(School_Id, Scores), y = Scores , fill = Gender)) +
  geom_boxplot(alpha = 0.7, color = "black") +
  labs(
    title = "Best and Worst Performed 5 Schools by Scores w.r.t Gender",
    x = "School ID", y = "Scores"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")+
  scale_fill_manual(values=c("lightpink","steelblue"))
```
*Interpretation*
Here school_ids 139, 93, 27 performs worse and overall boys perform better than girls while school_ids 65,5083,110 performs best and overall boys gives a better performance.

#Combined Genderwise plot of scores based on all variables 
```{r}

long_data <- cleaned_data %>%
  mutate(across(c(Gender, Parent_Educ, Parent_Support, Location, 
                  Book_Availability, Teacher_Shortage, Percent_Affluent,
                  Percent_Disadvantaged, Computer_Access, Year), 
                as.character)) %>%
  pivot_longer(
    cols = c(Parent_Educ, Parent_Support, Location, 
             Book_Availability, Teacher_Shortage,Percent_Affluent,
             Percent_Disadvantaged, Computer_Access, Year),
    names_to = "Variable",
    values_to = "Category"
  )


long_data <- long_data %>%
  mutate(Category = case_when(
    Variable == "Gender" ~ factor(Category, levels = c("GIRL", "BOY")),
    Variable == "Parent_Educ" ~ factor(Category, levels = c("less than ISCED2", "ISCED2", "ISCED3", "ISCED4", "ISCED6", "ISCED7 or 8")),
    Variable == "Parent_Support" ~ factor(Category, levels = c("LOW", "MEDIUM", "HIGH", "VERY HIGH")),
    Variable == "Location" ~ factor(Category, levels = c("Urban", "Urban_1","Suburban","Suburban_1",
                                             "Large Town/City", "Small Town/Village", "Remote Rural")),
    Variable == "Book_Availability" ~ factor(Category, levels = c("Few", "Moderate", "Good", "Very Good", "Excellent")),
    Variable == "Teacher_Shortage" ~ factor(Category, levels = c("None", "Low", "Moderate", "High")),
    Variable == "Percent_Affluent" ~ factor(Category, levels = c("Low", "Moderate", "High", "Excellent")),
    Variable == "Percent_Disadvantaged" ~ factor(Category, levels = c("Low", "Moderate", "High", "Severe")),
    Variable == "Computer_Access" ~ factor(Category, levels = c("NO", "YES")),
    Variable == "Year" ~ factor(Category, levels = c("2015", "2019")),
    TRUE ~ as.factor(Category)
  ))


ggplot(long_data, aes(x = Category, y = Scores, colour = Category)) +
  geom_boxplot() +
  facet_grid(Gender ~ Variable, scales = "free_x", switch = "x") +
  theme_minimal() +
  labs(
    title = "Gender-wise Distribution of Scores by Categorical Variables",
    x = "Category", y = "Scores"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


#Combined Genderwise plot of scores based on selected variables(significant) 
```{r}

long_data <- cleaned_data %>%
  mutate(across(c(Gender, Parent_Educ, Parent_Support, 
                  Book_Availability,Percent_Disadvantaged), 
                as.character)) %>%
  pivot_longer(
    cols = c(Parent_Educ, Parent_Support, 
             Book_Availability,Percent_Disadvantaged),
    names_to = "Variable",
    values_to = "Category"
  )


long_data <- long_data %>%
  mutate(Category = case_when(
    Variable == "Gender" ~ factor(Category, levels = c("GIRL", "BOY")),
    Variable == "Parent_Educ" ~ factor(Category, levels = c("less than ISCED2", "ISCED2", "ISCED3", "ISCED4", "ISCED6", "ISCED7 or 8")),
    Variable == "Parent_Support" ~ factor(Category, levels = c("LOW", "MEDIUM", "HIGH", "VERY HIGH")),
    Variable == "Book_Availability" ~ factor(Category, levels = c("Few", "Moderate", "Good", "Very Good", "Excellent")),
    Variable == "Percent_Disadvantaged" ~ factor(Category, levels = c("Low", "Moderate", "High", "Severe")),
    TRUE ~ as.factor(Category)
  ))


ggplot(long_data, aes(x = Category, y = Scores, colour = Category)) +
  geom_boxplot() +
  facet_grid(Gender ~ Variable,scales = "free_x", switch = "x") +
  theme_minimal() +
  labs(
    title = "Gender-wise Distribution of Scores based on Variables",
    x = "Category", y = "Scores"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

#Combined Scores Plot of selected variables by Location
```{r}

school_means <- cleaned_data %>%
  group_by(School_Id, Location) %>%
  summarise(Mean_Score = mean(Scores, na.rm = TRUE), .groups = "drop") %>%
  arrange(Mean_Score)


top_bottom_schools <- bind_rows(
  slice_head(school_means, n = 5),
  slice_tail(school_means, n = 5)
)


filtered_data <- cleaned_data %>%
  filter(School_Id %in% top_bottom_schools$School_Id)

filtered_data <- filtered_data %>%
  mutate(School_Location = paste(School_Id, "(", Location, ")", sep = ""))


filtered_data$School_Location <- as.factor(filtered_data$School_Location)

ggplot(filtered_data, aes(x = reorder(School_Location, Scores), y = Scores, fill = Location)) +
  geom_boxplot(alpha = 0.7, color = "black")+
  labs(
    title = "Top 5 and Bottom 5 Schools by Score and Location",
    x = "School (Location)",
    y = "Math Scores"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)+
  theme_minimal() +
  scale_fill_brewer(palette = "Spectral")+
  theme(legend.position = "bottom"))
```














