```{r}
#loading packages
library(dplyr)
library(tidyverse)
library(ggplot2)
```

```{r}
#loading dataframe
raw_data <- as.data.frame(irltim)
```

```{r}
view(cleaned_data)
```

#Data visualization

```{r}
ggplot(cleaned_data, aes(x = Scores)) +
  geom_histogram(binwidth = 20, fill = "red", color = "black", alpha = 0.7) +
  facet_wrap(~ Year) +
  ggtitle("Distribution of Scores by Year") +
  xlab("Scores") +
  ylab("Frequency") 
```

```{r}
ggplot(cleaned_data, aes(x = Year, y = Scores, fill = Gender)) +
  geom_boxplot(alpha = 0.6) + 
  ggtitle('Scores based on Gender w.r.t Year') +
  scale_fill_manual(values = c("pink", "blue")) 
```


```{r}
cleaned_data$Parent_Educ<- factor(cleaned_data$Parent_Educ,
                                  levels=c("less than ISCED2","ISCED2","ISCED3",
                                           "ISCED4","ISCED6","ISCED7 or 8"),
                                  ordered= TRUE)
cleaned_data$Year<-factor(cleaned_data$Year,levels=c("2015","2019"),ordered=TRUE)
ggplot(cleaned_data, aes(x = Parent_Educ, y = Scores, fill = Year)) +
  geom_boxplot(alpha = 0.6) + 
  ggtitle('Parent Education vs Scores') +
  scale_fill_brewer(palette = "Set3")
```


```{r}
cleaned_data$Year<-factor(cleaned_data$Year,levels=c("2015","2019"),ordered=TRUE)
ggplot(cleaned_data, aes(x = Parent_Support, y = Scores, fill = Year)) + 
  geom_boxplot(alpha=0.6) + 
  ggtitle('Parent Support vs Scores')

```
                    
```{r}
ggplot(cleaned_data, aes(x = Year, y = Scores, fill = Computer_Access)) +
  geom_boxplot(color = "black", alpha = 0.6) + 
  ggtitle('Computer Access vs Score based on Year') +
  scale_fill_manual(values = c("lightgreen", "red")) +
  theme_minimal()

```

```{r}
cleaned_data$Location<- factor(cleaned_data$Location,
                                  levels=c("Urban","Suburban","Large Town/City","Small Town/Village","Remote Rural"),
                                          ordered= TRUE)
ggplot(cleaned_data, aes(x = Gender, y = Scores, fill = Location)) + 
  geom_violin(color = "black", alpha = 0.6) + 
  ggtitle('Distribution of Scores by Location and Gender') + 
   scale_fill_manual(values=c("navyblue","brown","orange","lightgreen","lightblue")) 
+facet_wrap(~Year)
```
**Interpretation**

The shape of the violin plot represents density of scores within that specific category.The more wider the plots are the higher the density in that specific range. also narrower plots shows lower density which indicates only few students fall in that particular range.Also, if the scores are concentrated in the wider section of the violin it indicates less variability and scores that are more spread out indicates more variablity.
```{r}
cleaned_data$Book_Availability = factor(cleaned_data$Book_Availability,
                                        levels=c("Few","Moderate",
                                                  "Good","Very Good",
                                                           "Excellent"),
                                                            ordered=TRUE)
ggplot(cleaned_data, aes(x = Year, y = Scores, fill = Book_Availability)) + 
  geom_violin(color = "black", alpha = 0.6) + 
  ggtitle('Distribution of Scores by Book Availability and Gender') + 
  scale_fill_manual(values=c("lightblue","lightgreen","orange","brown","navyblue")) + 
  facet_wrap(~ Gender)
```
```{r}
cleaned_data$Teacher_Shortage =factor(cleaned_data$Teacher_Shortage,
                                      levels=c("None","Low","Moderate",
                                                       "High"),ordered=TRUE)
ggplot(cleaned_data, aes(x = Year, y = Scores, fill = Teacher_Shortage)) + 
  geom_violin(color = "black", alpha = 0.6) + 
  ggtitle('Distribution of Scores by Teacher Shortage Based on Years') + 
  scale_fill_brewer(palette = "Set3")+
  facet_wrap(~Gender)
```
```{r}
ggplot(cleaned_data, aes(x = Year, y = Scores, fill = Percent_Affluent)) + 
  geom_violin(alpha = 0.6) + 
  ggtitle('Distribution of Scores by Percent Affluent Based on Years') + 
  scale_fill_brewer(palette = "Set1")+
  facet_wrap(~Gender)
```
```{r}
ggplot(cleaned_data, aes(x = Year, y = Scores, fill = Percent_Disadvantaged)) + 
  geom_violin(alpha = 0.6) + 
  ggtitle('Distribution of Scores by Percent Disadvantaged Based on Years') + 
  scale_fill_brewer(palette = "Set1")+
  facet_wrap(~Gender)
```



















